Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} packages

### 2016-08-26 jss/Deveau.tex
Corrected master map
- Updated tables 2,3,4 and figures 1,2 for paper according to reviewer 2
comments
- Corrected table 1 with new "Create_master" function
- Split functions in three R files: data_handling, enrichment and plots

\begin{table}[h!]
 \centering
  \caption{ACSN maps included in the \pkg{ACSNMineR} package. Map: map name,
  Total: total number of gene symbols (HUGO) used to construct the map, Nb
  mod.: number of modules, Min: mimimum number of gene symbols in the modules,
  Max: maximum number of gene symbols in the modules, Mean: average number of
  gene sybols per module. N.B.: one gene symbol may be present in several
  modules of the map.} \label{tab:table1}
  \begin{tabular}{l c c c c c}
    \toprule
    Map & Total & Nb mod. & Min & Max & Mean\\
    \midrule
  ACSN global & 2243 & 63 & 2 &629& 83\\
  Survival  &1053&5  &208 &431 &328\\
  Apoptosis & 667&7 & 19& 382& 136\\
  EMT \& Cell motility &634 &9  &18 &629 &136\\
  DNA repair &345&21  &3 &171  &45\\
  Cell cycle &250&25  &2 &130  &19\\
  \bottomrule

@@

The example shown below is the simplest command that can be done to test a gene
list for over-representation on the six included ACSN maps. With the list of 12
genes mentionned above and a default p-value cutoff of $0.05$, we have a set of
36 maps or modules that are significantly enriched. The results are structured
as a data frame with nine columns displaying the module name, the module size,
the number of genes from the list in the module, the names of the genes that are
present in the module, the size of the reference universe, the number of genes
from the list that are present in the universe, the raw p-value, the p-value
corrected for multiple testing and the type of test performed. The module field
in the results data frame indicate the map name and the module name separated by
a column character. If a complete map is significantly enriched or depleted,
then only the map name is shown, without any module or column character. For
instance, the third line of the results object below concern the E2F1 module of
the CellCycle map.

@@

We previously computed the p-value of the \code{genes\_test} list with default
parameters. The number of modules which have a p-value below 0.05 was 36, that
can be compared to the 49 obtained without correction with the simple command
shown below (some of the results are displayed in table \ref{tab:table2}).

@@

\begin{widetable}[h!]
  \centering
  \caption{Overview of the results from enrichment analysis without correction.
  Module : name of the module. Mod. size: size of the module. Genes in module:
  genes from input which are found in the module. p-value: uncorrected p-value.
  Test : null hypothesis used, greater is synonym of enrichment.}
  \label{tab:table2}

  \begin{tabular}{l c c c c}
	\toprule
Module & Mod. size & Genes in module & p-value & Test\\
	\midrule
CellCycle & 242 & ATM ATR CHEK2   & 5.36e-07 & greater \\
          &   &  CREBBP KAT2B TFDP1   &          &         \\
          &   &  E2F1 EP300 HDAC1    &          &         \\
          &	  & GTF2H1 GTF2H2 GTF2H2B &	& \\
CellCycle: APOPTOSIS ENTRY & 10 & ATM ATR CHEK2 E2F1 & 3.49e-07 & greater \\
CellCycle: CYCLINB			& 7  & ATM 				  & 3.98e-02 & greater \\

@@

 represent_enrichment(enrichment = list(Corrected = results[1:6,], 
                             Uncorrected = results_uncorrected[1:6,]),
                               plot = "heatmap", scale = "reverselog", 
                               nrow = 1)
\end{example}

@@

In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
the calculation of p-values for enrichment or depletion of genes in biological
pathways. The package includes the six large-scale molecular maps and 55
functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment
can be calculated for those maps and modules with several options to play with,
but can also be calculated for other databases of molecular pathways, that can
be imported from GMT formated files.

>>>>

\begin{table}[h!]
 \centering
  \caption{ACSN maps included in the \pkg{ACSNMineR} package. Map: map name,
  Total: total number of gene symbols (HUGO) used to construct the map, Nb
  mod.: number of modules, Min: mimimum number of gene symbols in the modules,
  Max: maximum number of gene symbols in the modules, Mean: average number of
  gene sybols per module. N.B.: one gene symbol may be present in several
  modules of the map.} \label{tab:table1}
  \begin{tabular}{l c c c c c}
    \toprule
    Map & Total & Nb mod. & Min & Max & Mean\\
    \midrule
  ACSN global & 2239 & 67 & 2 &629& 79\\
  Survival  &1053&5  &208 &431 &328\\
  Apoptosis & 667&7 & 19& 382& 136\\
  EMT \& Cell motility &634 &9  &18 &629 &137\\
  DNA repair &345&21  &3 &171  &45\\
  Cell cycle &250&25  &2 &130  &20\\
  \bottomrule

@@

The example shown below is the simplest command that can be done to test a gene
list for over-representation on the six included ACSN maps. With the list of 12
genes mentionned above and a default p-value cutoff of $0.05$, we have a set of
8 maps or modules that are significantly enriched. The results are structured
as a data frame with nine columns displaying the module name, the module size,
the number of genes from the list in the module, the names of the genes that are
present in the module, the size of the reference universe, the number of genes
from the list that are present in the universe, the raw p-value, the p-value
corrected for multiple testing and the type of test performed. The module field
in the results data frame indicate the map name and the module name separated by
a column character. If a complete map is significantly enriched or depleted,
then only the map name is shown, without any module or column character. For
instance, the third line of the results object below concern the E2F1 module of
the CellCycle map.

@@

We previously computed the p-value of the \code{genes\_test} list with default
parameters. The number of modules which have a p-value below 0.05 was 8, that
can be compared to the 16 obtained without correction with the simple command
shown below (some of the results are displayed in table \ref{tab:table2}).

@@

\begin{widetable}[h!]
  \centering
  \caption{First rows of the results from enrichment analysis without correction.
  Module : name of the module. Mod. size: size of the module. Genes in module:
  genes from input which are found in the module. p-value: uncorrected p-value.
  Test : null hypothesis used, greater is synonym of enrichment.}
  \label{tab:table2}

  \begin{tabular}{l c c c c}
	\toprule
Module & Mod. size & Genes in module & p-value & Test\\
	\midrule
CellCycle & 242 & ATM ATR CHEK2 & $5.4 \times 10^{-7}$  & greater \\
		  &     & CREBBP TFDP1 E2F1 &  & \\
		  &     & EP300 HDAC1 KAT2B & & \\
		  &     & GTF2H1 GTF2H2 GTF2H2B &  &\\ 
  CellCycle:APOPTOSIS\_ENTRY & 10 & ATM ATR CHEK2 E2F1 & $3.5 \times 10^{-7}$  & greater \\ 
  CellCycle:CYCLINB & 7 & ATM & 0.04 & greater \\

@@

 represent_enrichment(enrichment = list(Corrected = results[1:6,], 
                             Uncorrected = results_uncorrected[1:6,]),
                               plot = "bar", scale = "reverselog", 
                               nrow = 1)
\end{example}

@@

In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
the calculation of p-values for enrichment or depletion of genes in biological
pathways. The package includes the six large-scale molecular maps and 67
functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment
can be calculated for those maps and modules with several options to play with,
but can also be calculated for other databases of molecular pathways, that can
be imported from GMT formated files.
